---
title: "Simulation"
author: "Michael Nsiah-Nimo"
format: html
execute:
  warning: false
---


# Warming Up

Most statistical software packages have several built-in distributions to handle the most common situations statisticians come across when simulating values.

Remember to set your seeds for these exercises so that you get the same answer each time the code is evaluated. 

## Exponential Distribution

1. Simulate 1000 draws from the exponential distribution $$f(x) = \lambda e^{-\lambda x},\;\;\;\; x\geq0, \text{ with } \lambda=3$$

```{r}
set.seed(7292)   # seed to reproduce the same results
n <- 1000
lambda <- 3

x <- rexp(n, rate = lambda)
head(x)
```



2. Create a density plot of your values. Add the PDF of the exponential(3) distribution to your plot in a different color. How do they compare?

```{r}
set.seed(7292) 

library(ggplot2)

# Create a data frame for the histogram
df <- data.frame(x = x)

# Create the plot
ggplot(df, aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 30, 
                 fill = "gray", 
                 color = "black") +
  stat_function(fun = dexp, 
                args = list(rate = lambda),
                color = "red", 
                linewidth = 1) +
  labs(title = "Simulated Exponential(3) vs True PDF",
       x = "x",
       y = "Density") +
  theme_minimal()
```

- Both declines exponentially as x increases
- The density estimate closely follows the theoretical PDF (red curve)
- The simulated sample behaves like the theoretical distribution.


3. How might  you use your 1000 draws to estimate the probability that $x > 1$? Provide an estimate using only your 1000 draws, and compare this estimate to one derived from `pexp` (R) or `scipy.stats.expon.cdf` (Python).

```{r}
set.seed(7292) 

# Monte Carlo estimate from sample
p_hat <- mean(x > 1)

# True value using CDF
p_true <- pexp(1, rate = lambda, lower.tail = FALSE)  # = P(X > 1)

p_hat
p_true
p_hat - p_true   # error

```


4. What is the average error from using a $n=1000$ simulation instead of a numerical calculation for the probability that $x>1$? Write a function to calculate the error from one simulation, and evaluate that function 100 times to calculate the expected error from using a simulation estimate rather than deterministic numerical integration.

```{r}
set.seed(7292)

lambda <- 3
threshold <- 1
p_true <- pexp(threshold, rate = lambda, lower.tail = FALSE)

one_sim_error <- function(n = 1000, lambda = 3, threshold = 1) {
  x <- rexp(n, rate = lambda)
  p_hat <- mean(x > threshold)
  error <- p_hat - p_true
  return(error)
}


errs <- replicate(100, one_sim_error(n = 1000))

mean(errs)          # average signed error
mean(abs(errs))     # average absolute error
sd(errs)            # variation across runs

```


## Gamma Distribution

The Gamma distribution is an exponential family, so it has MLEs that are unbiased and efficient estimators, if somewhat challenging to calculate in closed form. 

Luckily, there are functions in R and python to handle this:

- `fitdistr(...)` in the `MASS` package (R) will estimate the  parameters for a gamma distribution from sample data.

- `scipy.stats` distribution functions have a `.fit(data)` method that will estimate distributional parameters

In this problem, you will work with the gamma distribution with parameters $\alpha, \beta$ (note that R uses a different parameterization than Python) and PDF 

$$f(x | \alpha, \beta) = \frac{\beta^\alpha x^{\alpha-1} e^{-x}}{\Gamma(\alpha)}.$$

1. Write a function that will calculate the MLE $\hat\alpha,\hat\beta$ from a vector $x_{samp}$ of gamma-distributed samples. Your function should have required parameter $x$ and optional parameters $\alpha,\beta$ representing the true values. The function should return a list of $\hat\alpha, \hat\beta$ and, if $\alpha,\beta$ are provided, it should also return $\alpha_{err} = \alpha-\hat\alpha$ and $\beta_{err} = \beta-\hat\beta$. 

```{r}

library(MASS)

fit_gamma <- function(x, alpha = NULL, beta = NULL) {
  fit <- fitdistr(x, densfun = "gamma")
  
  alpha_hat <- fit$estimate[["shape"]]
  beta_hat  <- fit$estimate[["rate"]]  # rate = beta
  
  out <- list(alpha_hat = alpha_hat,
              beta_hat  = beta_hat)
  
  # If true parameters supplied, also return errors
  if (!is.null(alpha) && !is.null(beta)) {
    out$alpha_err <- alpha - alpha_hat
    out$beta_err  <- beta  - beta_hat
  }
  
  return(out)
}


# Example
#set.seed(7292)
#x_samp <- rgamma(100, shape = 2, rate = 4)  # true alpha=2, beta=4

#fit_gamma(x_samp, alpha = 2, beta = 4)
```


2. For a grid of $\alpha \times \beta=\{0.25, 0.50, 1.00, 2.00, 4.00\}\times\{0.25, 0.50, 1.00, 2.00, 4.00\}$, sample 100 values from the $\text{gamma}(\alpha,\beta)$ distribution. Provide the true parameter values and plot your error values in an appropriate plot. Take care to choose your plot mappings to support your discussion of the following: What trends (if any) do you notice in the estimation error?

```{r}
set.seed(7292)

alphas <- c(0.25, 0.5, 1, 2, 4)
betas  <- c(0.25, 0.5, 1, 2, 4)

grid <- expand.grid(alpha = alphas, beta = betas)


gamma_errors <- apply(grid, 1, function(row) {
  a <- as.numeric(row["alpha"])
  b <- as.numeric(row["beta"])
  
  x <- rgamma(100, shape = a, rate = b)
  fit <- fit_gamma(x, alpha = a, beta = b)
  
  c(alpha_err = fit$alpha_err,
    beta_err  = fit$beta_err)
})

gamma_errors <- as.data.frame(t(gamma_errors))
gamma_results <- cbind(grid, gamma_errors)

head(gamma_results)

```

Plot of errors
- Heatmap of alpha error

```{r}
library(ggplot2)

ggplot(gamma_results, aes(x = factor(alpha), y = factor(beta), fill = alpha_err)) +
  geom_tile() +
  scale_fill_gradient2(mid = 0) +
  labs(x = "alpha (shape)", y = "beta (rate)",
       fill = "alpha error",
       title = "Error in alpha_hat across (alpha, beta) grid")
```

- Heatmap of beta error
```{r}
library(ggplot2)

ggplot(gamma_results, aes(x = factor(alpha), y = factor(beta), fill = beta_err)) +
  geom_tile() +
  scale_fill_gradient2(mid = 0) +
  labs(x = "alpha (shape)", y = "beta (rate)",
       fill = "beta error",
       title = "Error in beta_hat across (alpha, beta) grid")
```


Trend:

- Errors are larger in magnitude for extreme parameter values (α very small or very large).


3. Now, let's vary the sample size. For $n=\{20, 30, 50, 100\}$, use your function to evaluate the error for $\alpha=1, \beta=2$. How does the error change as $n$ increases? Provide an appropriate plot as well as a description of the changes for increasing $n$. 

```{r}

set.seed(7292)

ns <- c(20, 30, 50, 100)

sim_list <- lapply(ns, function(n) {
  reps <- 500
  errs <- replicate(reps, {
    x <- rgamma(n, shape = 1, rate = 2)
    fit <- fit_gamma(x, alpha = 1, beta = 2)
    c(alpha_err = fit$alpha_err, beta_err = fit$beta_err)
  })
  errs <- as.data.frame(t(errs))
  errs$n <- n
  errs
})

gamma_n_results <- do.call(rbind, sim_list)

head(gamma_n_results)

```


- Plot of error vs n ( box plots)

```{r}

ggplot(gamma_n_results, aes(x = factor(n), y = alpha_err)) +
  geom_boxplot() +
  labs(x = "Sample size n", y = "alpha error",
       title = "Sampling distribution of alpha_hat error vs n")

ggplot(gamma_n_results, aes(x = factor(n), y = beta_err)) +
  geom_boxplot() +
  labs(x = "Sample size n", y = "beta error",
       title = "Sampling distribution of beta_hat error vs n")

```

As n increases:

- The center of the boxplots stays near 0, which implies the estimator is roughly unbiased
- The spread of the boxplots shrinks , implying that the estimates becomes more precise. This means both `alpha-hat` and `beta-hat` have smaller variability at larger `n`
